---
title: "CPRD_data_checking"
author: "Ric Campbell"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

## FOR INTERNAL PROJECT USE ONLY, PLEASE DO NOT SHARE BEYOND PROJECT TEAM

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(DataExplorer)
library(lubridate)
library(ggplot2)
library(dplyr)
library(kableExtra)
library(knitr)
library(readxl)
library(aurumpipeline)
library(arrow)
```

```{r set_up_patient_practice_data, echo = FALSE}

## Read in single file tables that want to check contain the same ids as each other, e.g. all 

  patients <- fread("D:/source_data/Delivery/Data/Primary care/TRAINS_Extract_Patient_001.txt",
                    colClasses = "character",
                    na.strings = "")
  
  practices <- fread("D:/source_data/Delivery/Data/Primary care/TRAINS_Extract_Practice_001.txt",
                    colClasses = "character",
                    na.strings = "")
  
  linkage_availability <- fread("D:/source_data/Delivery/Data/Linked data/linkage_eligibility_aurum.txt",
                                colClasses = "character",
                                na.strings = "")
  
  
  
  patient_imd <- fread("D:/source_data/Delivery/Data/Linked data/patient_imd2019_set22.txt",
                       colClasses = "character",
                       na.strings = "")
  
  randomisation <- fread("D:/reference_data/TRAINS Randomisation for CPRD.txt",
                         colClasses = "character",
                         na.strings = "")
  
  setnames(randomisation, make.names(colnames(randomisation), unique = TRUE))

  
## Change any dates to date types if needed for this
  
  patients[, cprd_ddate := as.Date(cprd_ddate, format = "%d/%m/%Y")]
  
## Patients without a imd index
  
  patients_no_imd <- patients[patid %chin% setdiff(patients[, patid], patient_imd[, patid])]
    
##  Practices that were randomised but not in practice/patient file  

  practices_randomised_no_data <- randomisation[CPRDPracticeId %chin% setdiff(randomisation[, CPRDPracticeId], practices[, pracid])]
  

## Ger number of patients per practice, and create field for groups of patients per practice
  
  number_patients_per_practice <- patients[, .N, by = pracid][order(N)]
  
  number_patients_per_practice[, patient_number_group := cut(N, c(0, 10, 100, 200, 300, 400, 500, 600), include.lowest = TRUE)]
  

## Calculate age at end of primary outcome measure point - 1st August to 30th September 2021
## So if born in September, they will have had birthday that year
## Easier to create fake date of birth to use lubridate for interval  
  
  patients[, created_dob := make_date(year = yob, month = mob, day = "01")]
  
  patients[, age := as.integer(interval(created_dob, as.Date("2021-09-30"))/dyears(1))]
  
  
## Merge practice randomisation to each patient to see split by arm, want to keep all patients
  
  patients_rand <- merge(patients,
                         randomisation[, .(CPRDPracticeId, Randomisation, Randomisation.Code)],
                         by.x = "pracid",
                         by.y = "CPRDPracticeId",
                         all = TRUE)
  
  stopifnot(patients[, .N] != patients_rand[, .N])

```

``` {R set_up_parquet_data, echo = FALSE}

## Create link to parquet files needed

  observation <- arrow::open_dataset("D:/source_data/parquet/observation",
                                        format = "parquet")

  drug <- arrow::open_dataset("D:/source_data/parquet/drug",
                               format = "parquet")
  
## Get the count of each code in observations in data table
  
  diag_counts <- data.table(observation %>%
                              select(medcodeid) %>%
                              map_batches(function(batch) {
                                batch %>%
                                  group_by(medcodeid) %>%
                                  summarise(N = n())
                              }) %>%
                            group_by(medcodeid) %>%
                            summarise(N = sum(N))
                            )
  
  drug_counts <- data.table(drug %>%
                              select(prodcodeid) %>%
                              map_batches (function(batch) {
                                batch %>%
                                  group_by(prodcodeid) %>%
                                  summarise(N = n())  
                              }) %>%
                            group_by(prodcodeid) %>%
                            summarise(N = sum(N))
                            )
                              
  
  
## Read in diagnosis code dictionary and merge with terms
  
  diag_terms <- fread("D:/reference_data/Asthma diagnosis codes 4 April.txt",
                      colClasses = "character",
                      na.strings = "")
  
  diag_terms_counts <- merge(diag_counts,
                             diag_terms[, .(MedCodeId, Term)],
                             by.x = "medcodeid",
                             by.y = "MedCodeId",
                             all = TRUE)

## Read in medicines code dictionary with terms
  
  drug_terms <- fread("D:/reference_data/Asthma related drug treatment codes 8 April.txt",
                      colClasses = "character",
                      na.string = "")
  
  drug_terms_count <- merge(drug_counts,
                            drug_terms[, .(ProdCodeId, ProductName)],
                            by.x = "prodcodeid",
                            by.y = "ProdCodeId",
                            all = TRUE)
  
  
```

```{R create_patient_desc, echo = FALSE}

## Take number of patients, practices, and patients per practice for report
  
  number_of_patients <- sprintf("%1.0f", patients[, .N])
  number_of_acceptable_patients <- sprintf("%1.0f", patients[acceptable == 1, .N])
  number_of_not_acceptable_patients <- sprintf("%1.0f", patients[acceptable == 2, .N])
  number_of_practices <- sprintf("%1.0f", practices[, .N])
  
  min_patients_in_practice <- sprintf("%1.0f", number_patients_per_practice[, min(N)])
  max_patients_in_practice <- sprintf("%1.0f", number_patients_per_practice[, max(N)])
  
  number_patients_no_imd <- sprintf("%1.0f", patients_no_imd[, .N])
  number_practices_randomised_no_data <- sprintf("%1.0f", practices_randomised_no_data[, .N])
  number_of_practices_no_region <- practices[is.na(region), .N]
  
  number_of_patients_lsoa <- sprintf("%1.0f", linkage_availability[lsoa_e == 1, .N])
  number_of_patients_no_lsoa <- sprintf("%1.0f", linkage_availability[is.na(lsoa_e) | lsoa_e == 0, .N])
  
  number_deaths_before_cutoff <- sprintf("%1.0f", patients[!is.na(cprd_ddate) & cprd_ddate < "2021-08-01", .N])
  
  
  
## Bind all these stats into a table
  
  pat_prac_desc <- rbind(number_of_patients, number_of_practices, min_patients_in_practice, max_patients_in_practice)

```


## Patient and Practice Counts

Table showing the number of patients and practices, and the min and max values of the number of patients per pratice in the dataset received

```{r pat_prac_desc, echo = FALSE}

  kable_styling(kable(pat_prac_desc, col.names = "value"),
                bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = TRUE)
```

## The number of patients by both age and gender in the dataset received

Only male and female gender have been included, less than 8 patients labelled 'I' (indeterminate) not includes

Email from CPRD stated *"However, once the patient turns 16 years of age the month is reset to 0. This change takes place at the practice or where a practice has stopped contributing month is reset in house by CPRD. This means that patients with a date of birth in 2006 and 2005 do not have a month of birth"*

Extract was created using the May CPRD Aurum release, assuming month field was updated on 31st May, 1st June 2006 would still be 15 and should have a month still available. There are no patients with a month of 0.

Numbers in the table have been rounded to the nearest 5 inline with stats disclosure rules

```{r, pat_age_gend, echo = FALSE}

  kable_styling(kable(patients[, .N, by = mob][order(as.integer(mob))][, N := round(N/5)*5]),
                bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = TRUE)

  kable_styling(kable(patients[, .N, by = yob][order(as.integer(yob))][, N := round(N/5)*5]),
                bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = TRUE)

  ggplot(patients[gender == 1 | gender == 2],
         aes(x = created_dob, fill = gender)) +
    geom_bar(position = "dodge") +
    ggtitle("Number of patients by age and gender") +
    scale_fill_discrete(labels = c("Male", "Female"))

```

## The number patients in each practice

N.B Numbers in the table have been rounded to the nearest 5 inline with stats discloure rules

```{r patient_per_prac, echo = FALSE}

  ggplot(number_patients_per_practice,
         aes(x = patient_number_group, y = N, fill = patient_number_group))+
    geom_bar(stat = "identity") + 
    ggtitle("Number of patients per pratice") + 
    scale_x_discrete(name = " Patients in practice")

## table of this as well (do kable if moving to RMarkdown)
  kable_styling(kable(number_patients_per_practice[, .N, by = patient_number_group][, N := round(N/5)*5]),
                bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = TRUE)
  
```  


## The number of patients in each arm of the trial

N.B Numbers in the table have been rounded to the nearest 5 inline with stats disclosure rules

```{r patients_rand, echo = FALSE}

  ggplot(patients_rand,
         aes(x = Randomisation, fill = Randomisation)) +
    geom_bar() + 
    ggtitle(("Number of patients per arm"))  

  kable_styling(kable(patients_rand[, .N, by = Randomisation][, N := round(N/5)*5]),
                bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = TRUE)
  
  
```

## Number of observations per term in the full dataset of observations provided

Only the diagnosis/terms that were selected in our code list are presented below, there are 39,766 other diagnosis present in the data.

N.B Numbers in the tables have been rounded to the nearest 5, with numbers under 8 being suppressed inline with stats disclosure rules.

```{r observation_counts, echo = FALSE}

  kable_styling(kable(diag_terms_counts[!is.na(Term)][order(N)][, N := round(N/5)*5][, N := as.character(N)][N == "5", N := "*"]),
                bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = TRUE)

```

## Number of drugs per product name in the full dataset of drugs issued provided

Only the drugs/terms that were selected in our code list are preesented below, there are 8,223 other drugs present in the data.

N.B Numbers in the tables have been rounded to the nearest 5, with numbers under 8 being suppressed inline with stats disclosure rules.

```{r drug_counts, echo = FALSE}

  kable_styling(kable(drug_terms_count[!is.na(ProductName) & ProductName != "NULL"][order(N)][, N := round(N/5)*5][, N := as.character(N)][N == "5", N := "*"]),
                bootstrap_options = c("striped", "hover", "condensed"), fixed_thead = TRUE)

```

## The number of observations, over time grouped by month, includes all observations in the provided data

```{r obs_graph, echo = FALSE}

## Graph, from parquet file, of all observations in data, over time grouped by month

  ggplot(observation %>%
           select(obsdate) %>%
           filter(obsdate != "obsdate" & !is.na(obsdate) & obsdate != "") %>%
           map_batches(function(batch) {
             batch %>%
              mutate(obsdate_month = paste0("01/", substr(obsdate, 4, 10))) %>%
              mutate(obsdate_month = as.Date(obsdate_month, format = "%d/%m/%Y")) %>%
              group_by(obsdate_month) %>%
              summarise(N = n())    
           }) %>%
           filter(obsdate_month >= "2004-01-01" & obsdate_month <= "2022-07-15") %>%
          group_by(obsdate_month) %>%
          summarise(N = sum(N)),
         aes(x = obsdate_month, y = N)) +
    geom_point() + 
    geom_smooth() + 
    scale_y_continuous(labels = scales::comma) + 
    ggtitle("Number of observations over time by month for all patients")


```

## The number of drug prescriptions, over time grouped by month, includes all observations in the provided data

```{r drug_graph, echo = FALSE}

## Graph, from parquet file, of all drugs in data, over time grouped by month

  ggplot(drug %>%
           select(issuedate) %>%
           filter(issuedate != "issuedate" & !is.na(issuedate) & issuedate != "") %>%
           map_batches(function(batch) {
             batch %>%
              mutate(issuedate_month = paste0("01/", substr(issuedate, 4, 10))) %>%
              mutate(issuedate_month = as.Date(issuedate_month, format = "%d/%m/%Y")) %>%
              group_by(issuedate_month) %>%
              summarise(N = n())
           }) %>%
           filter(issuedate_month >= "2004-01-01" & issuedate_month <= "2022-07-15") %>%
           group_by(issuedate_month) %>%
           summarise(N = sum(N)),
         aes(x = issuedate_month, y = N)) +
    geom_point() + 
    geom_smooth() + 
    scale_y_continuous(labels = scales::comma) + 
    ggtitle("Number of drug prescriptions over time by month for all patients")

```